Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-14 doc/thesis/methods_results.tex
Started writing intro for BA methods section

{\color{blue}\uline{ In this chapter, we will address the three research aims
of this thesis introduced in} \cref{sec:obj}. \uline{First, in}
\cref{sec:netabc}, \uline{we describe \software{netabc}, a computer program
that implements an \acrlong{ABC}-based algorithm to fit contact network models
to phylogenetic data. We also provide a justification for the use of \gls{ABC}
for this problem by arguing that the likelihood functions required to fit these
models by more conventional means are likely to be computationally intractable.
Second, in} \cref{sec:ba}, \uline{we perform a simulation study to investigate
the \acrlong{BA} network model, which uses a preferential attachment mechanism
to generate networks with the power law degree distributions observed in real
world social and sexual networks. We progress through two exploratory analyses
testing the identifiability of the model's parameters, and conclude by testing
\software{netabc}'s ability to recover the parameters from simulated
transmission trees. Third, in} \cref{sec:hiv}, \uline{we apply
\software{netabc} to fit the \gls{BA} model to six real world \gls{HIV}
datasets, with the understanding of the model parameters' identifiability
gained through the simulation experiments. We conclude the chapter with a
unified discussion of the three research aims, including interpretation of the
results of both the simulated and real data experiments, as well as an
examination of the limitations of our approach and opportunities for future
investigation.}}

\section{\software{Netabc}: a computer program for estimation of contact
network parameters with kernel-assisted ABC}
\label{sec:netabc}

\software{Netabc} is a computer program to perform statistical inference of
contact network parameters from an estimated transmission tree using
\gls{ABC}. {\color{blue}\uline{As discussed in }\cref{sec:obj}, \uline{the principal
statistical algorithm used by \software{netabc} is adaptive
\gls{ABC}-\gls{SMC}~\autocite{del2012adaptive}. In addition, there are two
supplementary components which are specific to the domain of phylogenetics and
contact networks: Gillespie simulation~\autocite{gillespie1976general}, to
simulate transmission trees on contact networks; and the tree
kernel~\autocite{poon2013mapping}, which is used as the distance function in
\gls{ABC} to compare transmission trees~\autocite{poon2015phylodynamic}} (see
\cref{sec:abc}).} We give a high-level overview of the program here, before
describing these components in detail. \software{Netabc} takes as input an
estimated transmission tree, which can be derived from a viral phylogeny by
rooting and time-scaling as described in \cref{subsec:phylodynamics} or
estimated by other methods~\autocite{cottam2008integrating,
jombart2011reconstructing, ypma2012unravelling, morelli2012bayesian,
didelot2014bayesian, hall2015epidemic}. We variously refer to this estimated
transmission tree as the observed tree, input tree, or true tree.

As described in \cref{sec:smc}, \software{netabc} keeps track of a population
of particles $x^{(k)}$, each of which contains particular parameter values
$\theta^{(k)}$ for the {\color{blue}\uline{contact network}} model we are
trying to fit {\color{blue}\uline{to the input tree}}. A small number of
contact networks $z^{(k)}$ are generated under the model for each particle, in
accordance with that particle's parameters. An epidemic is simulated over each
of these networks using Gillespie simulation, and by keeping track of its
progress, a transmission tree is obtained. Thus, each particle becomes
associated with several simulated transmission trees. These trees are compared
to the input tree using the tree kernel. Particles are weighted according to
the similarity of their associated simulated trees with the true tree, with
more similar trees receiving higher weights. The particles are iteratively
perturbed to explore the parameter space, and particles with simulated trees
too distant from the true tree are periodically dropped and resampled. Once a
convergence criterion is attained, the final set of particles is used as a
Monte Carlo approximation to the target distribution of \gls{ABC}, which is
assumed to resemble the posterior distribution on model parameters (see
\cref{sec:abc}). A graphical schematic of this algorithm is given in
\cref{fig:abcsmc}.

\begin{figure}
    \includegraphics{abc-smc.pdf}
    \caption[Graphical schematic of the ABC-SMC algorithm implemented in \software{netabc}.]{
      Graphical schematic of the \gls{ABC}-\gls{SMC} algorithm implemented in
      \software{netabc}. Particles are initially drawn from their prior
      distributions, making the initial population a Monte Carlo approximation
      to the prior. At each iteration, particles are perturbed, and a distance
      threshold around the true tree contracts. Particles are rejected, and
      eventually resampled, when all their associated simulated trees lie
      outside the threshold. As the algorithm progresses, the population
      smoothly approaches a Monte Carlo approximation of the \gls{ABC} target
      distribution, which is assumed to resemble the posterior.
    }
    \label{fig:abcsmc}
\end{figure}

\software{Netabc} is written in the \software{C} programming language. The
\software{igraph} library~\autocite{csardi2006igraph} is used to generate and
store contact networks and phylogenies. Judy arrays~\autocite{baskins2004judy}
are used for hash tables and dynamic programming matrices. The
\gls{GSL}~\autocite{gough2009gnu} is used to generate random draws from
probability distributions, and to perform the bisection step in the adaptive
\gls{ABC}-\gls{SMC} algorithm. Parallelization is implemented with \gls{POSIX}
threads~\autocite{barney2009posix}. In addition to the \software{netabc} binary
to perform kernel-assisted \gls{ABC}, we provide three additional stand-alone
utilities: \software{treekernel}, to calculate the tree kernel;
\software{nettree}, to simulate a transmission tree over a contact network; and
\software{treestat}, to compute various summary statistics of phylogenies. The
programs are freely available at \url{https://github.com/rmcclosk/netabc}.

To check that our implementation of Gillespie simulation was correct, we
reproduced Figure 1A of \textcite{leventhal2012inferring} (our
\cref{fig:leventhal}), which plots the imbalance of transmission trees
simulated over four network models at various levels of pathogen
transmissibility. Our implementation of adaptive \gls{ABC}-\gls{SMC} was tested
by applying it to the same mixture of Gaussians used by
\textcite{del2012adaptive} to demonstrate their method (originally used
by~\textcite{sisson2007sequential}). We were able to obtain a close
approximation to the function (see \cref{fig:smctest}), and attained the
stopping condition used by the authors in a comparable number of steps. To
check that the algorithm would converge to a bimodal distribution, we also
applied it to a mixture of two Gaussians with means $\pm$4 and variances 1. The
algorithm was able to recover both peaks (\cref{fig:smctest2}).

@@

{\color{blue}\uline{Using synthetic data,}} we investigated four parameters
related to the \gls{BA} contact network model, denoted \gls{N}, \gls{m},
\gls{alpha}, \gls{I} (see \cref{subsec:pa}). The first three of these are
parameters of the model itself, while \gls{I} is related to the simulation of
transmission trees over the network. However, we will refer to all four as
\gls{BA} parameters. \gls{N} denotes the total number of nodes in the network,
or equivalently, susceptible individuals in the population. When a node is
added to the network, \gls{m} new undirected edges are added incident to it,
and are attached to existing nodes of degree $k$ with probability proportional
to $k^\alpha + 1$ (\cref{subsec:pa}). To simulate transmission trees over a
\gls{BA} network, we allowed an epidemic to spread until \gls{I} nodes were
infected, and sampled a transmission tree at that time. The \gls{alpha}
parameter is unitless, while \gls{m} has units of edges or connections per
vertex, and \gls{N} and \gls{I} both have units of nodes or individuals.

@@

\begin{table}[ht]
  \centering
  \input{\tablepath/abc-expt}
  \caption[Variables used in grid search experiments]
  {
    Variables and \gls{BA} parameter values used for \gls{ABC} validation
    experiments. Trees were simulated under the test values, and
    kernel-assisted \gls{ABC} was used to re-estimate posterior distributions for the
    \gls{BA} parameters without training.
  }
  \label{tab:abcexpt}
\end{table}

@@

\Cref{fig:abcptm2} shows {\color{red}\sout{\gls{MAP}}}
{\color{blue}\uline{posterior mean}} point estimates of the BA model parameters
obtained with kernel-assisted ABC on simulated data. The estimates shown
correspond only to the simulations where the $m$ parameter was set to 2,
however the results for $m = 3$ and $m = 4$ were similar
(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% \gls{HPD}
intervals are given in \cref{tab:abchpd}.

@@

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Fri Jun 10 18:17:33 2016
\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \hline
Parameter & Estimate & Standard error & $p$-value \\ 
  \hline
(Intercept) & $2$ & $0.6$ & $0.01$ \\ 
  $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
  I & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
  m & $0.5$ & $0.2$ & $0.01$ \\ 
   \hline
\end{tabular}
\caption[Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters. GLM was fitted with a Gaussian distribution and inverse link function. Coefficients are interpretable as additive effects on the inverse of the mean error.} 
\label{tab:glmalpha}
\end{table}
% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Fri Jun 10 18:17:33 2016
\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \hline
Parameter & Estimate & Standard error & $p$-value \\ 
  \hline
(Intercept) & $0.004$ & $5\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
  $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
  I & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
  m & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
   \hline
\end{tabular}
\caption[Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters. GLM was fitted with a gaussian distribution and inverse link function. Coefficients are interpretable as additive effects on the inverse of the mean error.} 
\label{tab:glmI}
\end{table}

@@

{\color{red}\sout{
The accuracy of the parameter estimates obtained with kernel-assisted ABC
paralleled the results from the \gls{kSVR} classifier. Of the four parameters,
$\alpha$ was the most accurately estimated, with point estimates having a
median [IQR] absolute error of 
    0.11 
    [0.03 - 
    0.25].
The errors when the true value of $\alpha$ was zero were significantly greater
than those for the other values 
    (Wilcoxon rank-sum test, $p$ = $0$).
Errors in estimating $\alpha$ 
%also varied with the true value of $m$ just at
%the threshold of statistical significance
%    (one-way ANOVA, $p 
%    pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)$),
did not vary across the true values of $m$ or $I$ (both
one-way ANOVA).}

@@

\begin{table*}[ht]
  \centering
  \input{\tablepath/abc-hpd.tex}
  \caption[
      Average posterior mean point estimates and 95\% highest posterior density
      (HPD) interval widths for BA model parameter estimates obtained with
      kernel-assisted ABC.
  ]{
      Average posterior mean point estimates and 95\% highest posterior density
      (HPD) interval widths for BA model parameter estimates obtained with
      kernel-assisted ABC. Three transmission trees were simulated under each
      combination of the listed parameter values, and the parameters were
      estimated with kernel-assisted ABC without training.
  }
  \label{tab:abchpd}
\end{table*}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{abc-posterior-example}
  \vspace{6pt}
  \caption[
    Marginal posterior distributions of BA model parameters estimated
    with kernel-assisted ABC for a single simulated transmission tree.
  ]{
    Marginal posterior distributions of BA model parameters estimated
    with kernel-assisted ABC for a single simulated transmission tree. Dotted
    lines indicate true values, solid lines indicate posterior mean point
    estimates, and shaded areas show 95\% highest posterior density intervals.
  }
  \label{fig:abcex}
\end{figure}

@@

%<<abc_glm, include=FALSE>>=
%    source("global.R")
%    options(scipen=-1, digits=2)
%@
%
%<<alpha_glm, include=FALSE>>=
%    # alpha_error is influenced by alpha and m but not I
%    stopifnot(alpha.glm[Parameter == "alpha", min(p) < 0.05])
%    stopifnot(alpha.glm[Parameter == "m", min(p) < 0.05])
%    stopifnot(alpha.glm[Parameter == "I", min(p) > 0.05])
%
%    # alpha_error is correlated with true_alpha
%    alpha.test <- d[,cor.test(alpha_error, true_alpha, method="spearman")]
%    stopifnot(alpha.test$p.value < 0.05)
%
%    # alpha_error is not correlated with true_m
%    alpha.m.cor <- d[,cor.test(alpha_error, true_m)]
%    stopifnot(alpha.m.cor$p.value < 0.05)
%@
%
%We used \software{netabc} to estimate the parameters of the \gls{BA} model on
%simulated trees where the true parameter values were known. Point estimates for
%each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
%2. The results for the other values of \gls{m} were similar
%(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates
%of \gls{alpha} across all simulations was
%    d[,median(alpha_error)] 
%    [d[,quantile(alpha_error, 0.25)]-d[,quantile(alpha_error, 0.75)]].
%\Gls{GLM} analysis indicated that the true values of both \gls{alpha} and
%\gls{m} had significant effects on the error in estimated \gls{alpha}
%    ($p$ values $pp(alpha.glm[Parameter == "alpha", min(p)], eq=FALSE)$
%     $pp(alpha.glm[Parameter == "m", min(p)], eq=FALSE)$,
%     \cref{tab:glmalpha}),
%but the true value of \gls{I} did not. There was a significant negative
%correlation between the true value of \gls{alpha} and the error
%    (Spearman's $\rho$ = round(alpha.test$estimate, 2),
%     $p pp(alpha.test$p.value, eq=TRUE)$).
%There was a negative correlation between \gls{m} and the error in \gls{alpha}
%    (Spearman's $\rho$ = round(alpha.m.cor$estimate, 2),
%    $p pp(alpha.m.cor$p.value, eq=FALSE)$).
%
%<<I_glm, include=FALSE>>=
%    # I_error is influenced by alpha and I but not by m
%    stopifnot(I.glm[Parameter == "alpha", min(p) < 0.05])
%    stopifnot(I.glm[Parameter == "m", min(p) > 0.05])
%    stopifnot(I.glm[Parameter == "I", min(p) < 0.05])
%
%    # correlation between I_error and true_alpha
%    I.alpha.cor <- d[,cor.test(I_error, true_alpha, method="spearman")]
%    stopifnot(I.alpha.cor$p.value < 0.05)
%
%    # correlation between I_error and true_I
%    I.I.cor <- d[,cor.test(I_error, true_I, method="spearman")]
%    stopifnot(I.alpha.cor$p.value < 0.05)
%@
%
%The mean error in the estimated value of \gls{I} was
%    d[,round(median(I_error))] 
%    [d[,round(quantile(I_error, 0.25))]-d[,round(quantile(I_error, 0.75))]],
%an over-estimate of roughly a factor of 1.5
%(\cref{fig:abcptm2,fig:abcptm1,fig:abcptm3,fig:abcptm4}). \Gls{GLM} analysis
%indicated a relationship between the error in estimated \gls{I}, and the true
%values of \gls{alpha} and \gls{I}
%    ($p$ values $pp(I.glm[Parameter == "alpha", min(p)], eq=FALSE)$
%     $pp(alpha.glm[Parameter == "I", min(p)], eq=FALSE)$,
%     \cref{tab:glmalpha}),
%but not the true value of \gls{m}. There was a significant correlation between 
%the true value of \gls{alpha} and the error in estimated \gls{I}
%    (Spearman's $\rho$ = I.alpha.cor$estimate,
%     $p pp(I.alpha.cor$p.value, eq=TRUE)$),
%and between the true value of \gls{I} and the error in estimated \gls{I}
%    (Spearman's $\rho$ = I.I.cor$estimate,
%     $p pp(I.I.cor$p.value, eq=TRUE)$).
%
%<<m_glm, include=FALSE>>=
%    # effects of parameters on m_error
%    stopifnot(m.glm[Parameter == "m", min(p)] < 0.05)
%    stopifnot(m.glm[Parameter == "alpha", min(p)] > 0.05)
%    stopifnot(m.glm[Parameter == "I", min(p)] > 0.05)
%
%    # m was only estimated correctly in just over 20\% of simulations with m > 2
%    m.tbl <- prop.table(table(d[true_m > 1,m_error]))
%    stopifnot(m.tbl[1] > 0.2 & m.tbl[1] < 0.3)
%
%    # with m = 1, over 95% of simulations are right
%    m1.est <- d[true_m == 1, m_error]
%    stopifnot(sum(m1.est == 0) / length(m1.est) > 0.95)
%    stopifnot(sum(m1.est == 0) / length(m1.est) < 1)
%@
%
%\gls{GLM} analysis showed an effect of the true value of \gls{m} on the error
%in the estimated \gls{m}
%    ($p pp(m.glm[Parameter == "m", min(p)])$,
%     \cref{tab:glmm}).
%When the true value of \gls{m} was 1, the correct value was recovered by
%\software{netabc} in virtually every case
%    (sum(m1.est == 0) out of length(m1.est) simulations).
%However, when the true value of \gls{m} was 2 or higher, the correct value was
%recovered in only 
%    as.integer(m.tbl[1] * 100)\%
%of simulations, little better than random guessing. The \gls{GLM} analysis did
%not indicate any effects of the true parameter values on the error in estimated
%\gls{m} (\cref{tab:glmm}).
%
%<<N_glm, include=FALSE>>=
%    # alpha_error is influenced by alpha and m but not I
%    stopifnot(N.glm[Parameter == "alpha", min(p) < 0.05])
%    stopifnot(N.glm[Parameter == "I", min(p) < 0.05])
%    stopifnot(N.glm[Parameter == "m", min(p) > 0.05])
%
%    N.alpha.test <- d[,wilcox.test(.SD[true_alpha == 1.5, N_error], 
%                                   .SD[true_alpha < 1.5, N_error])]
%    stopifnot(N.alpha.test$p.value < 0.05)
%    N.I.test <- d[,wilcox.test(.SD[true_I == 3000, N_error], 
%                               .SD[true_I < 3000, N_error])]
%    stopifnot(N.I.test$p.value < 0.05)
%@
%
%Finally, the total number of nodes \gls{N} was consistently over-estimated by
%about a factor of two
%    (error d[,format(median(N_error), scientific=FALSE)] 
%    [d[,format(quantile(N_error, 0.25), scientific=FALSE)] - 
%     d[,format(quantile(N_error, 0.75), scientific=FALSE)]]).
%The fitted \gls{GLM} indicated that the true values of both \gls{alpha} and
%\gls{I} had an effect on the error in the estimated \gls{N}
%    ($p$-values $pp(N.glm[Parameter == "alpha", min(p)], eq=FALSE)$ and 
%     $pp(N.glm[Parameter == "I", min(p)], eq=FALSE)$,
%     \cref{tab:glmN}),
%but that the true value of \gls{m} did not. The error in the estimated \gls{N}
%when \gls{alpha} was equal to 1.5 was slightly lower than for other values of
%\gls{alpha}
%    (median [IQR] error rates 
%     d[true_alpha == 1.5, format(median(N_error), scientific=FALSE)] 
%    [d[true_alpha == 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
%     d[true_alpha == 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
%     for \gls{alpha} = 1.5 vs. 
%     d[true_alpha < 1.5, format(median(N_error), scientific=FALSE)] 
%    [d[true_alpha < 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
%     d[true_alpha < 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
%     for \gls{alpha} < 1.5),
%and this difference was statistically significant
%    (Wilcoxon rank-sum test, $p pp(N.alpha.test$p.value, eq=TRUE)$).
%Similarly, the error in the estimated \gls{N} when \gls{I} was equal to 3000 was
%lower than for other values of \gls{I}
%    (median [IQR] error rates 
%     d[true_I == 3000, format(median(N_error), scientific=FALSE)] 
%    [d[true_I == 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
%     d[true_I == 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
%     for \gls{I} = 3000 vs. 
%     d[true_alpha < 3000, format(median(N_error), scientific=FALSE)] 
%    [d[true_alpha < 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
%     d[true_alpha < 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
%     for \gls{I} < 3000),
%and this difference was statistically significant
%    (Wilcoxon rank-sum test, $p pp(N.I.test$p.value, eq=TRUE)$).
%
%\begin{figure}
%  \includegraphics{abc-point-estimate-m2}
%  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
%    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
%  {
%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
%    running \software{netabc} on simulated data. Values shown are for               
%    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
%    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
%    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
%    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
%  }
%  \label{fig:abcptm2}
%\end{figure}
%
%<<posterior_sims, echo=FALSE>>=
%    N <- 5000
%    replicate <- 0
%    lab <- NULL
%    for (m in c(2, 3, 4)) {
%    for (I in c(1000, 2000)) {
%    for (alpha in c(0, 0.5, 1, 1.5)) {
%        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
%    }
%    }
%    }
%    lab <- paste(lab, collapse=",")
%@
%
%\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
%distribution on the \gls{BA} parameters for one simulation. Equivalent plots
%for one replicate simulation with each combination of parameters can be found
%in \cref{lab}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
%narrow relative to the region of nonzero prior density, whereas the intervals
%for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
%estimates and 95\% \gls{HPD} intervals averaged over all simulations.
%
%\begin{figure}
%  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
%  \caption[Approximate marginal posterior distributions of BA model parameters
%      obtained by applying \textit{netabc} to a simulated transmission tree
%      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
%    {
%        Approximate marginal posterior distributions of BA model parameters
%        obtained by applying \textit{netabc} to a simulated transmission tree
%        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
%        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
%        highest posterior density intervals. $x$-axes indicate regions of
%        nonzero prior density.
%    }
%  \label{fig:abcex}
%\end{figure}
%
%
%\begin{table}
%    \centering
%    \input{\tablepath/abc-hpd}
%    \caption[
%        Maximum \textit{a priori} estimates and 95\% highest posterior density
%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
%        \software{netabc}, averaged over simulated transmission trees.
%    ]
%    {
%        Maximum \textit{a priori} estimates and 95\% highest posterior density

>>>>

\add{In this chapter, we will address the three research aims
of this thesis introduced in \cref{sec:obj}. First, in \cref{sec:netabc}, we
describe \software{netabc}, a computer program that implements an
\acrlong{ABC}-based algorithm to fit contact network models to phylogenetic
data. We also provide a justification for the use of \gls{ABC} for this problem
by arguing that the likelihood functions required to fit these models by more
conventional means are likely to be computationally intractable.  Second, in
\cref{sec:ba}, we perform a simulation study to investigate the \acrlong{BA}
network model, which uses a preferential attachment mechanism to generate
networks with the power law degree distributions observed in real world social
and sexual networks. We progress through two exploratory analyses testing the
identifiability of the model's parameters, and conclude by testing
\software{netabc}'s ability to recover the parameters from simulated
transmission trees. Third, in \cref{sec:hiv}, we apply \software{netabc} to fit
the \gls{BA} model to six real world \gls{HIV} datasets, with the understanding
of the model parameters' identifiability gained through the simulation
experiments. We conclude the chapter with a unified discussion of the three
research aims, including interpretation of the results of both the simulated
and real data experiments, as well as an examination of the limitations of our
approach and opportunities for future investigation.}

\section{\software{Netabc}: a computer program for estimation of contact
network parameters with ABC}
\label{sec:netabc}

\software{Netabc} is a computer program to perform statistical inference of
contact network parameters from an estimated transmission tree using
\gls{ABC}. \add{As discussed in \cref{sec:obj}, the principal statistical
algorithm used by \software{netabc} is adaptive
\gls{ABC}-\gls{SMC}~\autocite{del2012adaptive}. In addition, there are two
supplementary components which are specific to the domain of phylogenetics and
contact networks: Gillespie simulation~\autocite{gillespie1976general}, to
simulate transmission trees on contact networks; and the tree
kernel~\autocite{poon2013mapping}, which is used as the distance function in
\gls{ABC} to compare transmission trees~\autocite{poon2015phylodynamic} (see
\cref{sec:abc}).} We give a high-level overview of the program here, before
describing these components in detail. \software{Netabc} takes as input an
estimated transmission tree, which can be derived from a viral phylogeny by
rooting and time-scaling as described in \cref{subsec:phylodynamics} or
estimated by other methods~\autocite{cottam2008integrating,
jombart2011reconstructing, ypma2012unravelling, morelli2012bayesian,
didelot2014bayesian, hall2015epidemic}. We variously refer to this estimated
transmission tree as the observed tree, input tree, or true tree.

As described in \cref{sec:smc}, \software{netabc} keeps track of a population
of particles $x^{(k)}$ indexed by an integere $k$, each of which contains
particular parameter values $\theta^{(k)}$ for the \add{contact network} model
we are trying to fit \add{to the input tree}. A small number \gls{M} of contact
networks $z^{(k,i)}$, $1 \leq i \leq \gls{M}$, are generated under the model
for each particle in accordance with that particle's parameters. An epidemic is
simulated over each of these networks using Gillespie simulation, and by
keeping track of its progress, a transmission tree is obtained. Thus, each
particle becomes associated with several simulated transmission trees. These
trees are compared to the input tree using the tree kernel. Particles are
weighted according to the similarity of their associated simulated trees with
the true tree, with more similar trees receiving higher weights. The particles
are iteratively perturbed to explore the parameter space, and particles with
simulated trees too distant from the true tree are periodically dropped and
resampled. Once a convergence criterion is attained, the final set of particles
is used as a Monte Carlo approximation to the target distribution of \gls{ABC},
which is assumed to resemble the posterior distribution on model parameters
(see \cref{sec:abc}). A graphical schematic of this algorithm is shown in
\cref{fig:abcsmc}.

\begin{figure}
    \includegraphics{abc-smc.pdf}
    \caption[
        Graphical schematic of the ABC-SMC algorithm implemented in \software{netabc}.
    ]{
        Graphical schematic of the \gls{ABC}-\gls{SMC} algorithm implemented in
        \software{netabc}. Particles are initially drawn from their prior
        distributions, making the initial population a Monte Carlo
        approximation to the prior. At each iteration, particles are perturbed,
        and a distance threshold around the true tree contracts. Particles are
        rejected, and eventually resampled, when all their associated simulated
        trees lie outside the threshold. As the algorithm progresses, the
        population smoothly approaches a Monte Carlo approximation of the
        \gls{ABC} target distribution, which is assumed to resemble the
        posterior.
    }
    \label{fig:abcsmc}
\end{figure}

\software{Netabc} is written in the \software{C} programming language. The
\software{igraph} library~\autocite{csardi2006igraph} is used to generate and
store contact networks and phylogenies. Judy arrays~\autocite{baskins2004judy}
are used for hash tables and dynamic programming matrices. The
\gls{GSL}~\autocite{gough2009gnu} is used to generate random draws from
probability distributions, and to \del{perform the bisection step} \add{solve
for the next $\varepsilon$ by bisection} in the adaptive \gls{ABC}-\gls{SMC}
algorithm. Parallelization is implemented with \gls{POSIX}
threads~\autocite{barney2009posix}. In addition to the \software{netabc} binary
to perform \gls{ABC}, we provide three additional stand-alone utilities:
\software{treekernel}, to calculate the tree kernel; \software{nettree}, to
simulate a transmission tree over a contact network; and \software{treestat},
to compute various summary statistics of phylogenies. The programs are freely
available at \url{https://github.com/rmcclosk/netabc}.

To check that our implementation of Gillespie simulation was correct, we
reproduced Figure 1A of \textcite{leventhal2012inferring} (our
\cref{fig:leventhal}), which plots the imbalance of transmission trees
simulated over four network models at various levels of pathogen
transmissibility. Our implementation of adaptive \gls{ABC}-\gls{SMC} was tested
by applying it to the same mixture of Gaussians used by
\textcite{del2012adaptive} to demonstrate their method (originally used
by~\textcite{sisson2007sequential}). We were able to obtain a close
approximation to the function (see \cref{fig:smctest}), and attained the
stopping condition used by the authors in a comparable number of steps. To
check that the algorithm would converge to a bimodal distribution, we also
applied it to a mixture of two Gaussians with means $\pm4$ and variances 1. The
algorithm was able to recover both peaks (\cref{fig:smctest2}).

@@

\add{Using synthetic data,} we investigated four parameters related to the
\gls{BA} contact network model, 
denoted \gls{N}, \gls{m}, \gls{alpha}, \gls{I}
(see \cref{subsec:pa}). The first three of these are parameters of the model
itself, while \gls{I} is related to the simulation of transmission trees over
the network. However, we will refer to all four as \gls{BA} parameters. \gls{N}
denotes the total number of nodes in the network, or equivalently, susceptible
individuals in the population. When a node is added to the network, \gls{m} new
undirected edges are added incident to it, and are attached to existing nodes
of degree $d$ with probability proportional to $d^{\gls{alpha}} + 1$
(\cref{subsec:pa}). To simulate transmission trees over a \gls{BA} network, we
allowed an epidemic to spread until \gls{I} nodes were infected, and sampled a
transmission tree at that time. The \gls{alpha} parameter is unitless, while
\gls{m} has units of edges or connections per vertex, and \gls{N} and \gls{I}
both have units of nodes or individuals.

\end{document}

@@

\begin{table}[ht]
  \centering
  \input{\tablepath/abc-expt}
  \caption[Variables used in grid search experiments]
  {
    Variables and \gls{BA} parameter values used for \gls{ABC} validation
    experiments. Trees were simulated under the test values, and
    \gls{ABC} was used to re-estimate posterior distributions for the
    \gls{BA} parameters without training.
  }
  \label{tab:abcexpt}
\end{table}

@@

\Cref{fig:abcptm2} shows {\color{red}\sout{\gls{MAP}}}
{\color{blue}\uline{posterior mean}} point estimates of the BA model parameters
obtained with ABC on simulated data. The estimates shown
correspond only to the simulations where the $m$ parameter was set to 2,
however the results for $m = 3$ and $m = 4$ were similar
(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% \gls{HPD}
intervals are given in \cref{tab:abchpd}.

@@

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Tue Jun 14 16:56:16 2016
\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \hline
Parameter & Estimate & Standard error & $p$-value \\ 
  \hline
(Intercept) & $2$ & $0.6$ & $0.01$ \\ 
  $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
  I & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
  m & $0.5$ & $0.2$ & $0.01$ \\ 
   \hline
\end{tabular}
\caption[Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters. GLM was fitted with a Gaussian distribution and inverse link function. Coefficients are interpretable as additive effects on the inverse of the mean error.} 
\label{tab:glmalpha}
\end{table}
% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Tue Jun 14 16:56:16 2016
\begin{table}[ht]
\centering
\begin{tabular}{llll}
  \hline
Parameter & Estimate & Standard error & $p$-value \\ 
  \hline
(Intercept) & $0.004$ & $5\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
  $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
  I & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
  m & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
   \hline
\end{tabular}
\caption[Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters. GLM was fitted with a gaussian distribution and inverse link function. Coefficients are interpretable as additive effects on the inverse of the mean error.} 
\label{tab:glmI}
\end{table}

@@

{\color{red}\sout{
The accuracy of the parameter estimates obtained with ABC
paralleled the results from the \gls{kSVR} classifier. Of the four parameters,
$\alpha$ was the most accurately estimated, with point estimates having a
median [IQR] absolute error of 
    0.11 
    [0.03 - 
    0.25].
The errors when the true value of $\alpha$ was zero were significantly greater
than those for the other values 
    (Wilcoxon rank-sum test, $p$ = $0$).
Errors in estimating $\alpha$ 
%also varied with the true value of $m$ just at
%the threshold of statistical significance
%    (one-way ANOVA, $p 
%    pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)$),
did not vary across the true values of $m$ or $I$ (both
one-way ANOVA).}

@@

\begin{table*}[ht]
  \centering
  \input{\tablepath/abc-hpd.tex}
  \caption[
      Average posterior mean point estimates and 95\% highest posterior density
      (HPD) interval widths for BA model parameter estimates obtained with
      ABC.
  ]{
      Average posterior mean point estimates and 95\% highest posterior density
      (HPD) interval widths for BA model parameter estimates obtained with
      ABC. Three transmission trees were simulated under each
      combination of the listed parameter values, and the parameters were
      estimated with ABC without training.
  }
  \label{tab:abchpd}
\end{table*}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{abc-posterior-example}
  \vspace{6pt}
  \caption[
    Marginal posterior distributions of BA model parameters estimated
    with ABC for a single simulated transmission tree.
  ]{
    Marginal posterior distributions of BA model parameters estimated
    with ABC for a single simulated transmission tree. Dotted
    lines indicate true values, solid lines indicate posterior mean point
    estimates, and shaded areas show 95\% highest posterior density intervals.
  }
  \label{fig:abcex}
\end{figure}

@@
